---
title: "Phase 1p SNPs in ResolveOME and call breakpoints"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = FALSE}
# rmarkdown::render('src/resolveome/nf-resolveome/03_phase_snps.Rmd', output_file = 'phase_snps.html', output_dir = 'out/phase_snps/')

# chunk options
cache_path <- "out/phase_snps/phase_snps_cache/"
knitr::opts_knit$set(root.dir = "../../..")
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  cache.path = cache_path,
  fig.align = "center"
)
```

```{r setup2, include = FALSE}
# create cache path
dir.create(cache_path, showWarnings = FALSE)

# libraries
library(knitr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(rtracklayer)
library(GenomicRanges)
library(slider)
library(patchwork)

# function: plot baf
plot_baf <- function(p_dat) {
  # prep data
  p_dat2 <-
    p_dat %>%
    filter(total_depth > 4) %>%
    dplyr::mutate(mut_baf = 1 - mut_vaf,
                  chr = factor(sub("^chr", "", chr),
                               levels = c(as.character(1:22), "X", "Y")),
                  `pos (Mb)` = pos / 1e6) %>%
    tidyr::pivot_longer(cols = c("mut_vaf", "mut_baf"), names_to = "vaf_type")

  # prep phased data
  p_dat_phased <-
    p_dat2 %>%
    filter(!is.na(hap_A_vaf) & vaf_type == "mut_vaf")

  # plot
  p_dat2 %>%
    ggplot(aes(x = `pos (Mb)`, y = value)) +
    # add line at vaf = 0.5
    geom_hline(yintercept = 0.5, colour = "red") +
    # add baf points
    geom_point(size = 0.8, alpha = 0.2) +
    # add phased points
    geom_point(data = p_dat_phased, aes(y = hap_A_vaf), colour = "#e03ba9",
               size = 1) +
    geom_point(data = p_dat_phased, aes(y = hap_B_vaf), colour = "#00ccff",
               size = 1) +
    # add line at TNFRSF14
    geom_vline(xintercept = 2487078 / 1e6, colour = "green") +
    # increase x axis breaks
    scale_x_continuous(expand = c(0, 0),
                       breaks = pretty(p_dat2$`pos (Mb)`, n = 20)) +
    ggh4x::facet_grid2(. ~ chr, scales = "free_x", space = "free_x") +
    theme_classic() +
    theme(panel.grid.major = element_line(),
          panel.grid.minor.x = element_line(),
          strip.background = element_rect(color = "grey", fill = NA,
                                          linewidth = 0, linetype = "solid")) +
    # rotate x axis text using guides
    guides(x = guide_axis(angle = -90))
}
```

The longest 1p LOH event is observed in cells `plate3_wellA2`, affecting the
entire 1p arm, and is therefore expected to be homozygous for all SNPs. We can
therefore phase all SNPs in PD63118 chr1p. Then, I will look for genic SNPs, as 
these will be most likely to be picked up in the snRNAseq.

We get the cell IDs with 1p LOH.

```{r load_metadata}
loh_cell_ids <-
  "data/resolveome/manual_inspection/PD63118.tsv" %>%
  readr::read_tsv() %>%
  filter((!chr_dropout | is.na(chr_dropout)) &
         (!suspected_doublet | is.na(suspected_doublet)) &
         loh_1p) %>%
  pull(cell_id)
```

First, we get all SNPs on 1p in the LOH cells.

```{r load_plate3_wellA2_snps, eval = F}
loh_snps <-
  "out/resolveome/nf-resolveome/dna/PD63118/genotyping/snps/PD63118_genotyped_snps.tsv" %>%
  readr::read_tsv() %>%
  mutate(cell_id = gsub("_dna.*", "", id)) %>%
  filter(cell_id %in% loh_cell_ids, chr == 1) %>%
  split(.$id)

# save
saveRDS(loh_snps, file.path(cache_path, "loh_genotyped_chr1_snps.rds"))
```

```{r reload_snps}
loh_snps <- readRDS(file.path(cache_path, "loh_genotyped_chr1_snps.rds"))
```

Next, we phase the SNPs on 1p in `plate3_wellA2`.

```{r phase_plate3_wellA2_snps}
plate3_wellA2 <-
  loh_snps$plate3_wellA2_dna_run49882 %>%
  filter(pos < 121500000, total_depth > 4) %>%
  # assign haplotypes
  mutate(hap_A_type = case_when(mut_vaf > 0.9 ~ "alt",
                                mut_vaf < 0.1 ~ "ref"),
         hap_A_vaf = case_when(mut_vaf > 0.9 ~ mut_vaf,
                               mut_vaf < 0.1 ~ 1 - mut_vaf),
         hap_B_type = case_when(mut_vaf > 0.9 ~ "ref",
                                  mut_vaf < 0.1 ~ "alt"),
         hap_B_vaf = case_when(mut_vaf > 0.9 ~ 1 - mut_vaf,
                               mut_vaf < 0.1 ~ mut_vaf)) %>%
  right_join(loh_snps$plate3_wellA2_dna_run49882)
```

Now, we visualise the phased SNPs for `plate3_wellA2` in a BAF plot.

```{r plot_baf_plate3_wellA2_snps, fig.width = 25, fig.height = 5}
plot_baf(plate3_wellA2)
```

## Phased BAF plots

Now we plot this phasing in all other LOH cell.

```{r plot_bafs, fig.width = 25, fig.height = 5}
# phase snps
phased_loh_snps <-
  purrr::map(loh_snps, function(loh_snps_i) {
    loh_snps_i %>%
      left_join(
        plate3_wellA2 %>%
          distinct(chr, pos, ref, alt, hap_A_type, hap_B_type) %>%
          filter(!is.na(hap_A_type))) %>%
      mutate(hap_A_vaf = case_when(hap_A_type == "alt" ~ mut_vaf,
                                   hap_A_type == "ref" ~ 1 - mut_vaf),
             hap_B_vaf = case_when(hap_B_type == "alt" ~ mut_vaf,
                                   hap_B_type == "ref" ~ 1 - mut_vaf))
  })

# save phased snps
saveRDS(phased_loh_snps, file.path(cache_path, "phased_loh_genotyped_chr1_snps.rds"))

# save phased baf plots
purrr::walk2(names(phased_loh_snps), phased_loh_snps, function(id_i, i) {
  p <- plot_baf(i) + ggtitle(id_i)
  print(p)
})
```

## Breakpoints

Next, we attempt to identify the breakpoints in each case.

```{r id_breakpoints, fig.height = 10, fig.width = 25}
# define whether you care about 1 or -1 (based on hap A or B present)
# find the first position at which the rolling average delta is > -0.9 or < 0.9
# take the snp one before that
# plot a red dotted line and record the position of the breakpoint
breakpoints_and_haps <- list()
purrr::walk2(names(phased_loh_snps), phased_loh_snps, function(id_i, i) {

  print(id_i)

  # which haplotype is present based on the first 100 phased snps?
  present_hap <-
    i %>%
    filter(!is.na(hap_A_vaf)) %>%
    head(100) %>%
    summarise(A = mean(hap_A_vaf),
              B = mean(hap_B_vaf)) %>%
    tidyr::pivot_longer(cols = everything()) %>%
    filter(value == max(value)) %>%
    pull(name)

  # get rolling average vaf delta of phased snps only
  p_dat <-
    i %>%
    # only look at phased snps with high coverage
    filter(!is.na(hap_A_vaf), total_depth >= 10) %>%
    # get rolling average vaf delta
    arrange(pos) %>%
    mutate(`pos (Mb)` = pos / 1e6,
           vaf_delta = case_when(present_hap == "A" ~ hap_A_vaf - hap_B_vaf,
                                 present_hap == "B" ~ hap_B_vaf - hap_A_vaf),
           rolling_median_vaf_delta = slide_dbl(vaf_delta, median,
                                                .before = 100, .after = 100),
           n = row_number())

  # get breakpoint
  breakpoint_n <-
    p_dat %>%
    filter(rolling_median_vaf_delta < 0.9) %>%
    head(1) %>%
    # get the previous snp
    {.$n - 1}
  if (length(breakpoint_n) == 0) {
    breakpoint_n <- nrow(p_dat)
  }
  breakpoint <- p_dat %>% filter(n == breakpoint_n) %>% pull(pos)

  # plot
  p1 <- plot_baf(i) + ggtitle(id_i)
  p2 <-
    p_dat %>%
    ggplot(aes(x = `pos (Mb)`, y = rolling_median_vaf_delta)) +
    geom_line() +
    geom_vline(xintercept = breakpoint / 1e6, colour = "red",
               linetype = "dashed") +
    # make y limits 0 and 1
    scale_y_continuous(limits = c(-1, 1)) +
    scale_x_continuous(limits = c(min(i$pos / 1e6), max(i$pos / 1e6)),
                       expand = c(0, 0)) +
    ggtitle(paste0("breakpoint = ", breakpoint))
  p <- p1 / p2
  print(p)

  # save breakpoint and present haplotype
  breakpoints_and_haps[[id_i]] <<-
    tibble::tibble(breakpoint = breakpoint, present_haplotype = present_hap)

})
```

Now, we plot the breakpoints for each cell.

```{r plot_breakpoints, fig.height = 8}
# convert breakpoints to tibble
breakpoints_and_haps <-
  bind_rows(breakpoints_and_haps, .id = "id")

# load the established clades
clades <-
  "data/resolveome/shared_clades/shared_clades.tsv" %>%
  readr::read_tsv()

# plot breakpoints
breakpoints_and_haps %>%
  mutate(cell_id = gsub("_dna.*", "", id)) %>%
  left_join(clades) %>%
  ggplot(aes(x = reorder(cell_id, -breakpoint), y = breakpoint / 1e6,
             fill = clade)) +
  geom_col() +
  geom_text(aes(label = present_haplotype), nudge_y = 1.5) +
  geom_hline(yintercept = 2487078 / 1e6, colour = "green") +
  theme_classic() +
  labs(x = "", y = "chr1p breakpoint (Mb)") +
  guides(x = guide_axis(angle = -90))

# save breakpoints
breakpoints %>%
  readr::write_tsv(file.path(cache_path, "breakpoints.tsv"))
```